---
title: "Spatio-temporal KDE"
format: html
editor: visual
---

## Loading R packages

```{r}
pacman::p_load(tidyverse, sf, spatstat, sparr, lubridate)
```

## Data Preparation

Code chunk below is used import mpsz.rds into RStudio and then to extract Ang Mo Kio planning area boundary from *mpsz* sf data.frame.

```{r}
mpsz_am <- read_rds("../data/rds/mpsz.rds") %>%
  filter(PLN_AREA_C == "AM")
```

Next, we will import the *origin_sf.rds* into RStudio

```{r}
origin_sf <- read_rds("../data/rds/origin_sf.rds")
```

The code below is used to extract grab trips originated from Ang Mo Kio planning area.

```{r}
origin_am <- st_intersection(origin_sf, mpsz_am)
```

Because **ppp** object only accept numerical or character as mark. The code chunk below is used to convert data type of *starttime* to numeric.

```{r}
origin_am$tt <- as.numeric(origin_am$starttime)
```

The code chunk below is used to remove the unwanted fields from *origin_am* sf data.frame. This is because `as.ppp()` only need the mark field and geometry field from the input sf data.frame.

```{r}
origin_am <- origin_am %>%
  select(tt)
```

### Creating ppp

The code chunk below is used to derive a **ppp** object called *origin_am_ppp* from *origin_am sf* data.frame.

```{r}
origin_am_ppp <- as.ppp(origin_am)
origin_am_ppp
```

The code chunk below is used to check the output is in the correct object class.

```{r}
summary(origin_am_ppp)
```

Next, we will check if there are duplicated point events by using the code chunk below.

```{r}
any(duplicated(origin_am_ppp))
```

### Including Owin object

Next, `as.owin()` is used to convert *mpsz_am* into an **owin** object.

```{r}
am_owin <- as.owin(mpsz_am)
```

Next, `class()` is used to confirm if the output is indded an owin object.

```{r}
class(am_owin)
```

The code chunk below is used to combine *origin_am_ppp* and *am_owin* objects into one.

```{r}
origin_am_ppp <- origin_am_ppp[am_owin]
summary(origin_am_ppp)
```

As a good practice, `plot()` is used to plot *oringin_am_ppp* so that we can examine the correctness of the output object.

```{r}
plot(origin_am_ppp)
```

## Computing Spatio-temporal KDE

In the code chunk below, [`spattemp.density()`](https://tilmandavies.github.io/sparr/reference/spattemp.density.html) of **sparr** package is used to compute the spatial-temporal KDE by using origin_am_ppp as the input.

```{r}
kde_origin_am <- spattemp.density(origin_am_ppp,
                                  tres=128)
summary(kde_origin_am)
```

### Plotting the spatio-temporal KDE object

```{r}
tims <- c(1555000000,1555100000,1555200000,
          1555300000,1555400000,1555500000,
          1555600000,1555700000)
par(mfcol=c(3,3))
for(i in tims){ 
  plot(kde_origin_am, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at time", as_datetime(i)))
}
```
