---
pagetitle: KDE Analysis | Capstone
title: Temporal KDE Analysis using sparr package
---

# Getting Started

The code chunk below uses `p_load()` function of pacman package to check if the required packages have been installed on the computer. If they are, the packages will be launched.

The packages used are:

-   sf package is used for importing, managing, and processing geospatial data.
-   tidyverse package for aspatial data wrangling.
-   spatstat package to perform 1st-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.
-   tmap package is used to visualize geospatial data on maps.

```{r}
options(scipen = 999999)

pacman::p_load(sf, tidyverse, spatstat, tmap, sparr, lubridate, magick)
```

In the code chunk below, the [cleaned data files](DataPrep.qmd) are loaded into R.

```{r}
# Load base map
mpsz = st_read(dsn = "../data/geospatial", 
               layer = "MPSZ-2019") %>%
  st_transform(crs = 3414)

# Load rds file
pickup_sf <- st_as_sf(read_rds("../data/rds/pickup_sf.rds")) %>%
  st_intersection(mpsz) %>%
  mutate(pingtimestamp = as.numeric(pingtimestamp)) %>%
  select(pingtimestamp)

dropoff_sf <- st_as_sf(read_rds("../data/rds/dropoff_sf.rds")) %>%
  st_intersection(mpsz) %>%
  mutate(pingtimestamp = as.numeric(pingtimestamp)) %>%
  select(pingtimestamp)
```

# Convert Pickup and Dropoff location sf dataframes into spatstat’s ppp object format

Convert the sf dataframe to a ppp object using `as.ppp()` function from the spatstat package. The result is a marked planar point pattern. To change a marked planar point pattern to just a planar point pattern, simply remove the marks associated with each point using `marks(pickup_ppp) <- NULL`.

```{r}
pickup_ppp <- as.ppp(pickup_sf)

plot(pickup_ppp)
```

```{r}
dropoff_ppp <- as.ppp(dropoff_sf)

plot(dropoff_ppp)
```

# Create observational window (owin)

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In spatstat, an object called owin is specially designed to represent this polygonal region. The code chunk below is used to create a border of Singapore's land area and convert it into an owin object of spatstat package using `as.owin()`.

```{r}
# Create border of Singapore's land area
mpsz_border <- st_cast(mpsz)

# Convert the resulting sf object to an owin object
mpsz_owin <- as.owin(mpsz_border)

plot(mpsz_owin)
```

# Combine point events object and owin object

In the code chunks below, pick-up and drop-off events located within the observational window are extracted.

```{r}
pickup_owin = pickup_ppp[mpsz_owin]

plot(pickup_owin)
```

```{r}
dropoff_owin = dropoff_ppp[mpsz_owin]

plot(dropoff_owin)
```

# TKDE for Pick-up locations using adaptive bandwidth

```{r}
#| eval: false
pickup_scaled <- rescale(pickup_owin, 1000, "km")

bw = bw.abram(pickup_scaled)

pickup_tkde <- spattemp.density(pickup_scaled, h = bw, tres = 128)
```

## Plot gif

```{r}
#| eval: false
summary(as_datetime(pickup_scaled$marks))
```

```{r}
#| eval: false
# Define your dates (assuming dates is a vector of date values)
dates <- as.numeric(seq(min(as_datetime(pickup_scaled$marks)), 
                        max(as_datetime(pickup_scaled$marks)), 
                        by = "3 hours"))

# Create a temporary directory to save the images
dir.create("west_pickup")

# Generate and save each plot
for (i in dates) {
  # Define the file name
  file_name <- paste0("west_pickup/plot_", as.character(i), ".png")
  
  # Open a PNG device
  png(filename = file_name)
  
  # Get the day of the week
  day_of_week <- wday(as_datetime(i), label = TRUE, abbr = TRUE, week_start = 1)
  
  # Create the plot
  plot(pickup_tkde, i, 
       override.par = FALSE, 
       fix.range = TRUE, 
       main = paste("KDE at time", day_of_week, as_datetime(i)))
  
  # Close the PNG device
  dev.off()
}

```

```{r}
#| eval: false
# List the image files
image_files <- list.files("west_pickup", full.names = TRUE, pattern = "*.png")

# Read the images into a list
image_list <- lapply(image_files, image_read)

# Combine the images into an animation
animation <- image_animate(image_join(image_list), fps = 2)

# Save the animation as a GIF
image_write(animation, "west_pickup.gif")

```

## Plot spatial and temporal outputs

```{r}
#| eval: false
png(paste0("Spatial margin - Pick-up.png"), width=500, height=350)
plot(pickup_tkde$spatial.z,box=FALSE,main="Spatial margin\n Pick-up")
dev.off()
png(paste0("Temporal margin - Pick-up.png"), width=750, height=500)
plot(pickup_tkde$temporal.z,xlim=pickup_tkde$tlim,xaxs="i",main="Temporal margin\n Pick-up")
dev.off()
```

# TKDE for Drop-off locations using adaptive bandwidth

```{r}
#| eval: false
dropoff_scaled <- rescale(dropoff_owin, 1000, "km")

bw = bw.abram(dropoff_scaled)

dropoff_tkde <- spattemp.density(dropoff_scaled, h = bw, tres = 128)
```

## Plot gif

```{r}
#| eval: false
summary(as_datetime(dropoff_scaled$marks))
```

```{r}
#| eval: false
# Define your dates (assuming dates is a vector of date values)
dates <- as.numeric(seq(min(as_datetime(dropoff_scaled$marks)), 
                        max(as_datetime(dropoff_scaled$marks)), 
                        by = "3 hours"))

# Create a temporary directory to save the images
dir.create("west_dropoff")

# Generate and save each plot
for (i in dates) {
  # Define the file name
  file_name <- paste0("west_dropoff/plot_", as.character(i), ".png")
  
  # Open a PNG device
  png(filename = file_name)
  
  # Get the day of the week
  day_of_week <- wday(as_datetime(i), label = TRUE, abbr = TRUE, week_start = 1)
  
  # Create the plot
  plot(dropoff_tkde, i, 
       override.par = FALSE, 
       fix.range = TRUE, 
       main = paste("KDE at time", day_of_week, as_datetime(i)))
  
  # Close the PNG device
  dev.off()
}

```

```{r}
#| eval: false
# List the image files
image_files <- list.files("west_dropoff", full.names = TRUE, pattern = "*.png")

# Read the images into a list
image_list <- lapply(image_files, image_read)

# Combine the images into an animation
animation <- image_animate(image_join(image_list), fps = 2)

# Save the animation as a GIF
image_write(animation, "west_dropoff.gif")

```

## Plot spatial and temporal outputs

```{r}
#| eval: false
png(paste0("Spatial margin - Drop-off.png"), width=500, height=350)
plot(dropoff_tkde$spatial.z,box=FALSE,main="Spatial margin\n Drop-off")
dev.off()
png(paste0("Temporal margin - Drop-off.png"), width=750, height=500)
plot(dropoff_tkde$temporal.z,xlim=dropoff_tkde$tlim,xaxs="i",main="Temporal margin\n Drop-off")
dev.off()
```