---
pagetitle: KDE Analysis | Capstone
title: Temporal KDE Analysis using sparr package
---

# Getting Started

The code chunk below uses `p_load()` function of pacman package to check if the required packages have been installed on the computer. If they are, the packages will be launched.

The packages used are:

-   sf package is used for importing, managing, and processing geospatial data.
-   tidyverse package for aspatial data wrangling.
-   spatstat package to perform 1st-order spatial point patterns analysis and derive kernel density estimation (KDE) layer.
-   tmap package is used to visualize geospatial data on maps.

```{r}
options(digits = 10)

pacman::p_load(sf, tidyverse, spatstat, tmap, sparr, lubridate, magick)
```

In the code chunk below, the [cleaned data files](DataPrep.qmd) are loaded into R.

```{r}
# Load base map
mpsz = st_read(dsn = "../data/geospatial", 
               layer = "MPSZ-2019") %>%
  st_transform(crs = 3414) %>%
  filter(!PLN_AREA_N %in% c("WESTERN ISLANDS", "NORTH-EASTERN ISLANDS"))

# Load rds file
pickup_sf <- st_as_sf(read_rds("../data/rds/pickup_sf.rds")) %>%
  # convert to hours
  mutate(pingtimestamp = as.numeric(pingtimestamp) / 3600) %>%
  select(pingtimestamp)

dropoff_sf <- st_as_sf(read_rds("../data/rds/dropoff_sf.rds")) %>%
  #convert to hours
  mutate(pingtimestamp = as.numeric(pingtimestamp) / 3600) %>%
  select(pingtimestamp)
```

# Convert Pickup and Dropoff location sf dataframes into spatstat’s ppp object format

Convert the sf dataframe to a ppp object using `as.ppp()` function from the spatstat package. The result is a marked planar point pattern. To change a marked planar point pattern to just a planar point pattern, simply remove the marks associated with each point using `marks(pickup_ppp) <- NULL`.

```{r}
pickup_ppp <- as.ppp(pickup_sf)

plot(pickup_ppp)
```

```{r}
dropoff_ppp <- as.ppp(dropoff_sf)

plot(dropoff_ppp)
```

# Create observational window (owin)

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In spatstat, an object called owin is specially designed to represent this polygonal region. The code chunk below is used to create a border of Singapore's land area and convert it into an owin object of spatstat package using `as.owin()`.

```{r}
# Create border of Singapore's land area
mpsz_border <- st_cast(mpsz %>%
                         filter(REGION_N == "CENTRAL REGION"))

# Convert the resulting sf object to an owin object
mpsz_owin <- as.owin(mpsz_border)

plot(mpsz_owin)
```

# Combine point events object and owin object

In the code chunks below, pick-up and drop-off events located within the observational window are extracted.

```{r}
pickup_owin = pickup_ppp[mpsz_owin]

plot(pickup_owin)
```

```{r}
dropoff_owin = dropoff_ppp[mpsz_owin]

plot(dropoff_owin)
```

# TKDE for Pick-up locations

```{r}
#| eval: false
pickup_scaled <- rescale(pickup_owin, 1000, "km")

h_bw <- bw.ppl(pickup_scaled)

lambda_bw <- bw.ucv(marks(pickup_scaled))

pickup_tkde <- spattemp.density(pickup_scaled, h = h_bw, lambda = lambda_bw, tres = 84)
```

## Plot spatial outputs

```{r}
#| eval: false
# Define date values
earliest_date <- min((pickup_scaled$marks * 3600))
latest_date <- max(as.numeric(pickup_scaled$marks * 3600))
dates <- seq(earliest_date, latest_date, by = 3600 * 6)

# Create directory to save images
dir_name <- "pickup_central"
dir.create(dir_name)

# Generate and save each plot
for (i in dates) {
  # Define the file name
  file_name <- paste0(dir_name, "/plot_", as.character(i), ".png")
  
  # Open a PNG device
  png(filename = file_name)
  
  # Get the day of the week
  day_of_week <- wday(as_datetime(i), label = TRUE, abbr = TRUE, week_start = 1)
  
  # Create the plot
  plot(pickup_tkde, i/3600, 
       override.par = FALSE, 
       fix.range = TRUE, 
       main = paste(day_of_week, as_datetime(i)))
  
  # Close the PNG device
  dev.off()
}
```

## Plot temporal outputs

```{r}
#| eval: false
temporal_density <- pickup_tkde$temporal.z

# Create a data frame for plotting
plot_data <- data.frame(
  Date = as_datetime(temporal_density$x * 3600),
  Density = temporal_density$y
)

# Plot the data using ggplot2
temporal_margin <- ggplot(plot_data, aes(x = Date, y = Density)) +
  geom_line() +
  scale_x_datetime(date_labels = "%b-%d", date_breaks = "1 day", minor_breaks = NULL) +
  theme_minimal() +
  labs(title = "Temporal margin\n Pick-up Events in Central Region",
       x = "Date",
       y = "Events per square kilometre per hour)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_line(color = "gray80"),
        panel.grid.minor = element_line(color = "gray90"),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.title = element_text(hjust = 0.5))

temporal_margin

ggsave('pickup_temp_central.png', temporal_margin, width = 12, height = 5.5)
```

# TKDE for Drop-off locations

```{r}
#| eval: false
dropoff_scaled <- rescale(dropoff_owin, 1000, "km")

h_bw <- bw.ppl(dropoff_scaled)

lambda_bw <- bw.ucv(marks(dropoff_scaled))

dropoff_tkde <- spattemp.density(dropoff_scaled, h = h_bw, lambda = lambda_bw, tres = 84)
```

## Plot spatial outputs

```{r}
#| eval: false
# Define date values
earliest_date <- min((dropoff_scaled$marks * 3600))
latest_date <- max(as.numeric(dropoff_scaled$marks * 3600))
dates <- seq(earliest_date, latest_date, by = 3600 * 6)

# Create directory to save images
dir_name <- "dropoff_central"
dir.create(dir_name)

# Generate and save each plot
for (i in dates) {
  # Define the file name
  file_name <- paste0(dir_name, "/plot_", as.character(i), ".png")
  
  # Open a PNG device
  png(filename = file_name)
  
  # Get the day of the week
  day_of_week <- wday(as_datetime(i), label = TRUE, abbr = TRUE, week_start = 1)
  
  # Create the plot
  plot(dropoff_tkde, i/3600, 
       override.par = FALSE, 
       fix.range = TRUE, 
       main = paste(day_of_week, as_datetime(i)))
  
  # Close the PNG device
  dev.off()
}
```

## Plot temporal outputs

```{r}
#| eval: false
temporal_density <- dropoff_tkde$temporal.z

# Create a data frame for plotting
plot_data <- data.frame(
  Date = as_datetime(temporal_density$x * 3600),
  Density = temporal_density$y
)

# Plot the data using ggplot2
temporal_margin <- ggplot(plot_data, aes(x = Date, y = Density)) +
  geom_line() +
  scale_x_datetime(date_labels = "%b-%d", date_breaks = "1 day", minor_breaks = NULL) +
  theme_minimal() +
  labs(title = "Temporal margin\n Drop-off Events in Central Region",
       x = "Date",
       y = "Events per square kilometre per hour)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.major = element_line(color = "gray80"),
        panel.grid.minor = element_line(color = "gray90"),
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.title = element_text(hjust = 0.5))

temporal_margin

ggsave('dropoff_temp_central.png', temporal_margin, width = 12, height = 5.5)
```

